{{< include preamble.Rmd >}}

# Social Factors

> The Social Determinants of Health (SDOH) are the upstream non-health factors
> that “impact a wide range of health, functioning and quality of life outcomes.”1
> For public health, this is as preventive as it gets. When considering these
> upstream factors in the work of a public health entity such as the Connecticut
> Department of Public Health (CT DPH), we can more effectively inform the public
> and policymakers so we can all live better lives. As an agency whose mission
> declares that the equal enjoyment of a person’s highest attainable standard of
> health is a human right, we must also examine the conditions that contribute to
> “avoidable differences in health among specific population groups that result
> from cumulative social disadvantages.”[@ctdph-health-disparities] 
> More specifically, we apply an equity lens to ascertain which populations 
> are being most negatively impacted.[^social-factors-1]

[^social-factors-1]: We can do footnotes on highlight, that 
also show up at the bottom of the page

Look references on highlight - go find that same
reference easily in the original document. It's not clickable, and the pdf
is 400 pages long so it's had to find the reference page, and it takes a while
to load. So much friction, now references are easier to work
with.

## Education

Economic factors such as poverty and unemployment can lead
to unhealthy living conditions. Yet education can provide
individuals with foundational knowledge, life skills, and social
and psychological supports to make healthier choices. Therefore,
quality education and higher educational attainment can be a
protective factor that can advance more equitable outcomes[-@education2014].
It has been demonstrated that individuals without a high school
diploma have higher incidences of risk behaviors and other
adverse health outcomes; and earn less money, which can limit
access to resources and healthy environments[-@barhii2015].

### Early Education

Experiences and education within the first five years of life can
shape one’s health trajectory across the lifespan. Early education
and care programs can be protective against social and economic
challenges and narrow inequitable gaps in health outcomes.5
Participating in these programs are also associated with higher
educational attainment, better eating habits, increased use of
preventive healthcare services, and lower rates of child injuries,
child abuse/maltreatment, teen pregnancy, depression, use of
tobacco or other drugs, and arrests and incarceration.

As noted in @fig-pct-enroll, the rate of Pre-K enrollment for 4-year old
children in state-funded preschool programs in Connecticut has made sizable
gains over the past 5 years; however, enrollment among 3-year old children has
remained fairly stable over the past decade and thus far peaked at 10% in 2016.

---

> CONNECTICUT RANKS NINTHS IN THE US FOR EARLY EDUCATION
> SPENDING PER CHILD[^social-factors-2]

---

[^social-factors-2]: Another footnote :)

- Connecticut has three state-funded
pre-kindergarten programs: School Readiness
Program, Child Day Care Contracts, and
Smart Start[^social-factors-3]

- In 2018, 14,585 children were enrolled in state
pre-kindergarten programs[^social-factors-3]

- On average, our state spent $7,612 per child
enrolled; this reflects a 30% drop in average per
child expenditure since 2011[^social-factors-3]

- According to most recent estimates available
(2013–2014 school year), children enrolled
in CT early childhood and pre-kindergarten
programs were:[^social-factors-4]  

  - 50% are non-Hispanic White, 26% are Hispanic/Latino,
15% are non-Hispanic Black or African American, and
5% are non-Hispanic Asian

  - 23% are students with disabilities served under the
 Individuals with Disabilities Education Act
 
  - 2% are English Language Learners (those who speak
English less than “very well;” for more information about
Limited English Proficient section)

[^social-factors-3]: Friedman-Krauss, A.H., et al. (2019). The State of
Preschool 2018: State Preschool Yearbook. National Institute for Early Education
Research. Retrieved from
http://nieer.org/wp-content/uploads/2019/05/YB2018_Executive-SummaryR.pdf.

[^social-factors-4]: Civil Rights Data Collection, 2013–2014 State and National
Estimates: Total Enrollment in Early Childhood and Pre-K.

---

```{r}
#| label: fig-pct-enroll
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Quick Example Plot without CT Colors

library(ggplot2)
library(dplyr)

four_yo <- c(9,10,13,15,14,16,16,11,13,13,13,13,14,23,24,30,30)
three_yo <- c(3,3,2,2,3,4,5,8,7,7,8,6,6,9,10,8.5,9)

prek_enrollment <- 
  dplyr::tibble(
    pct_enroll = c(four_yo, three_yo),
    year = c(2002:2018, 2002:2018),
    age = c(rep("4 year olds", 17), rep("3 year olds", 17))
  )

basic_plot <- 
  ggplot(prek_enrollment, aes(year, pct_enroll)) +
  geom_point(aes(col = age)) +
  geom_line(aes(col = age)) +
  theme_classic() +
  labs(
    title = 
    "Percentage of children enrolled in early childhood and Pre-K programs",
    subtitle = "by age group CT, 2002-2018",
    col = "",
    x = "",
    y = "Percentage of Population Enrolled"
  )

basic_plot

#TODO add guides
#TODO add proper colors
#TODO add source caption
#TODO more years on x axis
#TODO remove y axis line
  
  
  
```


```{r}
#| label: fig-pct-enroll-ct
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Same Plot with CT Colors and random source @ct-rdl-2024


blue <- list_color_hex_codes$list_ct_primary_colors$primary
orange <- list_color_hex_codes$list_ct_secondary_colors$secondary

basic_plot +
   scale_color_manual(values = c(blue, orange)) +
  theme_ct_vertical() +
  theme_poppins_bold_all() +
  theme(
    title = element_text(size = 8)
  )
 

```


### Education Attainment

{{< lipsum 3 >}}

## Neighborhood and Built Environment

### Access to Health Foods

{{< lipsum 1 >}}

### Food Insecurity

{{< lipsum 2 >}}
